Efficient simulation of individual-based population models

### 2024-04-02 article.qmd
update

---
title: "Efficient simulation of individual-based population models"
subtitle: "The `R` package `IBMPopSim`"
author:
  - name: Daphné Giorgi 
    email: daphne.giorgi@sorbonne-universite.fr
    url: https://perso.lpsm.paris/~giorgi/
    affiliations:
      - name: Sorbonne Université, CNRS
        department: Laboratoire de Probabilités, Statistique et Modélisation
        url: https://www.lpsm.paris/
  - name: Vincent Lemaire 
    corresponding: true
    email: vincent.lemaire@sorbonne-universite.fr
    url: https://perso.lpsm.paris/~vlemaire/
    orcid: 0000-0002-0433-7722
    affiliations:
      - name: Sorbonne Université, CNRS
        department: Laboratoire de Probabilités, Statistique et Modélisation
        url: https://www.lpsm.paris/
  - name: Sarah Kaakai
    corresponding: true
    email: sarah.kaakai@univ-lemans.fr
    url: https://perso.univ-lemans.fr/~skaakai/
    affiliations:
      - name: Le Mans Université
        department: Laboratoire Manceau de Mathématiques
        url: https://lmm.univ-lemans.fr/fr/
      - name: Ecole Polytechnique, CNRS
        department: Centre de Mathématiques Appliquées
        url: https://cmap.ip-paris.fr/ 
date: last-modified
date-modified: last-modified
description: |
    This document provides a full description of the Stochastic Individual-Based Models (IBMs) that can be implemented in the IBMPopSim package. A unified mathematical and simulation framework is given, with a detailed description of the simulation algorithm. Examples of applications for the package are also provided, showing the performance and flexibility of IBMPopSim. 
abstract: >+
    The `R` Package `IBMPopSim` aims to simulate the random evolution of heterogeneous populations using stochastic Individual-Based Models (IBMs). The package enables users to simulate population evolution, in which individuals are characterized by their age and some characteristics, and the population is modified by different types of events, including births/arrivals, death/exit events, or changes of characteristics. The frequency at which an event can occur to an individual can depend on their age and characteristics, but also on the characteristics of other individuals (interactions). Such models have a wide range of applications in fields including  actuarial science, biology, ecology or epidemiology. `IBMPopSim` overcomes the limitations of time-consuming IBMs simulations by implementing new efficient algorithms  based on thinning methods, which are compiled using the `Rcpp` package while providing a user-friendly interface.
keywords: [Individual-based models, stochastic simulation, population dynamics, Poisson measures, thinning method, actuarial science, insurance portfolio simulation]
citation:
  type: article-journal
  container-title: "Computo"
  doi: "xxxx"
  url: https://computo.sfds.asso.fr/template-computo-quarto
  publisher: "Société Française de Statistique"
  issn: "2824-7795"
bibliography: references.bib
github-user: VincentLemaire
repo: "ibmpopsim_computo"
draft: true # set to false once the build is running
published: false # will be set to true once accepted
format:
  computo-html: default
  computo-pdf: default
crossref:
    cnj-title: Assumption
    cnj-prefix: Assumption 
execute:
    eval: true
    echo: true
    warning: false
---

@@

```{r, echo=FALSE, fig.width=8}
#| label: fig-birth-size
#| fig-cap: "Evolution of birth size"
#| fig-pos: 'h'

>>>>

---
title: "Efficient simulation of individual-based population models"
subtitle: "The `R` package `IBMPopSim`"
author:
  - name: Daphné Giorgi 
    corresponding: true
    email: daphne.giorgi@sorbonne-universite.fr
    url: https://perso.lpsm.paris/~giorgi/
    affiliations:
      - name: Sorbonne Université, CNRS
        department: Laboratoire de Probabilités, Statistique et Modélisation
        url: https://www.lpsm.paris/
  - name: Sarah Kaakai
    corresponding: true
    email: sarah.kaakai@univ-lemans.fr
    url: https://perso.univ-lemans.fr/~skaakai/
    affiliations:
      - name: Le Mans Université
        department: Laboratoire Manceau de Mathématiques
        url: https://lmm.univ-lemans.fr/fr/
      - name: Ecole Polytechnique, CNRS
        department: Centre de Mathématiques Appliquées
        url: https://cmap.ip-paris.fr/ 
  - name: Vincent Lemaire 
    corresponding: true
    email: vincent.lemaire@sorbonne-universite.fr
    url: https://perso.lpsm.paris/~vlemaire/
    orcid: 0000-0002-0433-7722
    affiliations:
      - name: Sorbonne Université, CNRS
        department: Laboratoire de Probabilités, Statistique et Modélisation
        url: https://www.lpsm.paris/
date: last-modified
date-modified: last-modified
description: |
    This document provides a full description of the Stochastic Individual-Based Models (IBMs) that can be implemented in the IBMPopSim package. A unified mathematical and simulation framework is given, with a detailed description of the simulation algorithm. Examples of applications for the package are also provided, showing the performance and flexibility of IBMPopSim. 
abstract: >+
    The `R` Package `IBMPopSim` aims to simulate the random evolution of heterogeneous populations using stochastic Individual-Based Models (IBMs). The package enables users to simulate population evolution, in which individuals are characterized by their age and some characteristics, and the population is modified by different types of events, including births/arrivals, death/exit events, or changes of characteristics. The frequency at which an event can occur to an individual can depend on their age and characteristics, but also on the characteristics of other individuals (interactions). Such models have a wide range of applications in fields including  actuarial science, biology, ecology or epidemiology. `IBMPopSim` overcomes the limitations of time-consuming IBMs simulations by implementing new efficient algorithms  based on thinning methods, which are compiled using the `Rcpp` package while providing a user-friendly interface.
keywords: [Individual-based models, stochastic simulation, population dynamics, Poisson measures, thinning method, actuarial science, insurance portfolio simulation]
citation:
  type: article-journal
  container-title: "Computo"
  doi: "xxxx"
  url: https://computo.sfds.asso.fr/template-computo-quarto
  publisher: "Société Française de Statistique"
  issn: "2824-7795"
bibliography: references.bib
github-user: VincentLemaire
repo: "ibmpopsim_computo"
draft: true # set to false once the build is running
published: false # will be set to true once accepted
format:
    computo-html: default
    computo-pdf: default
crossref:
    cnj-title: Assumption
    cnj-prefix: Assumption 
execute:
    eval: true
    echo: true
    warning: false
---

@@

```{r, echo=FALSE, fig.width=8}
#| label: fig-birth-size
#| fig-cap: "Evolution of birth size"
#| fig-pos: 'h'
#| fig-format: png
